Continous time Markov chain models; \\ extension, simulation and visualisation

### 2015-02-11 Response.tex

%%%%%%%%%%%%%%%%
%---DOCUMENT---%
%%%%%%%%%%%%%%%%
\begin{document}

% consistent edits between manuscript and rebuttal letter
\input{thesis.xtr}

\setcounter{page}{1}

\newcommand{\myTitle}{Continuous-time Markov chain models for pathogen phylodynamics: }
\newcommand{\mySubtitle}{model extension, simulation and visualisation} % produkcja dystrybucja masturbacja :)

\bigskip
\medskip
\noindent{\Large 
Response to the reviewer comments to: \myTitle \mySubtitle.
}
\bigskip

@@

\begin{itemize}

% 1
\item {
{\it
\textbf{
The thesis starts with a very nice introduction, which I think is very insightful for the reader who is not familiar with phylogenetic and/or Bayesian inference. 
In spite of this, I believe that the readability of the later chapters would be further improved if the introduction:
}% END: textbf
}% END: it
}% END: item

\begin{itemize}

% 1a
\item {
{\it
\textbf{
Briefly described the evolutionary models, such as GY94, that are later used.
}% END: textbf
}% END: it
}% END: item
\\
Corrected, there is now a \emph{Markov models of codon evolution} subsection in the introductory chapter, which discusses most popular codon model implementations:

\begin{quote}
\myeditsvonea
\end{quote}

% 1b
\item {
{\it
\textbf{
Briefly described how to accommodate time-stamped sequences, which is now considered in the analyses in Chapter 2, but not really discussed.
}% END: textbf
}% END: it
}% END: item
\\
Corrected, I added a subsection which describes the methods used in BEAST to accomodate time-stamped sequences as a source of calibration for trees, when such information is availiable:

\begin{quote}
\myeditsvoneb
\end{quote}

% 1c
\item {
{\it
\textbf{
Give a bit more detail on phylogeography, e.g. how do the models look like and what is the connection between phylogeography and epoch models (some connection is described in Chapter 2, but remains a bit vague).
}% END: textbf
}% END: it
}% END: item
\\
I added a figure and additional explanation, which should make the subsequent chapters clearer now. 
These changes start with:

\begin{quote}
\myeditsvonec
\end{quote}

@@

\begin{quote}
\myeditsvthree
\end{quote}

% 4
\item {
{\it
\textbf{
Expression (1.12) is confusing in that indices i and j appear in the lefthand side, but not in the righthand side. 
Likewise in expression (1.13).
}% END: textbf
}% END: it
}% END: item
\\
Corrected.

% 5
\item {
{\it
\textbf{
P20, 2nd display: an equality sign is missing on the 2nd line.
}% END: textbf
}% END: it
}% END: item
\\
Corrected.

% 6
\item {
{\it
\textbf{
I would find it useful to see a brief discussion on the plausibility of (1.22).
}% END: textbf
}% END: it
}% END: item
\\
% FB: Does he mean plausibility from a biological perspective or statistical? It's a necessary condition for the existence of stationary dist.
% Pl: I see it as questioning the plausibility in a biological sense. We now already argue that it is an assumption taken for computational reasons, but at the end of that paragraph we can explicitly state 'Realistic substitution processes are however unlikely to be time-reversible and procedures have also been developed to compute phylogenetic likelihoods under non-reversible models (cite Boussau and Gouy, 2006, see below)'. ...perhaps add in something more from that paper.... and end with 'We note that we use such non-reversible models in our discrete phytogeographic applications (cite Edwards et al., Current Biology, 2011).
% http://citeseerx.ist.psu.edu/viewdoc/download?doi=10.1.1.324.620&rep=rep1&type=pdf
Please see my changes which explain how the assumptions of stationarity and homogeneity of stochastic process governing the evolution connect to each other

with the assumption of

and which start with:

\begin{quote}
\myeditsvsixone
\end{quote}

In addition to explaining the implications of this assumption from biological perspective, we also briefly introduce some of the recent attempts to relax this assumption, both in the Bayesian phylogeographic framework and more bradly in the context of phylogenetic applications:

\begin{quote}
\myeditsvsixtwo
\end{quote}

### 2015-02-11 chapters/chapter01.tex
response

As $t$ goes to infinity the probability that the site is in some state $j$ is non-zero and does not depend on the starting state $i$.
This implies that the chain will always display the same long-term behavior, which is why $\mathbf{\Pi}=\{\pi_{i},\ i\in\mathcal{E}\}$ is called a \emph{stationary distribution} (also referred to as the \emph{equilibrium distribution} or \emph{equilibrium frequencies}).
We can look at the elements of $\mathbf{\Pi}$ as the proportion of time spent in state $j\in\mathcal{E}$ after the Markov chain has run for infinitely long time, that does not depend upon the initial state $i$ of the process.
In other words, if we sample the initial state from the stationary distribution and then let the process run for some sufficiently long time $t$, then the distribution of the final state will equal the stationary distribution. 
From the biological perspective stationarity means that the probability distribution of a certain state (e.g. nucleotide character) is constant over the whole course of evolution; the probability to draw a certain state is the same whenever the sampling is done.
A process can be stationary and not be homogeneous, for example by varying substitution rates on the tree branches (see \cite{Yang2002} and the discussion in Subsection~\ref{sub:dpp}).
However the converse is not true, and non-stationarity induces non-homogeneity, as the stochastic process of evolution depends on the stationary distribution.

@@

Realistic substitution processes are however unlikely to be time-reversible, especially since abandoning stationarity and homogeneity assumptions implies abandoning reversibility of the evolutionary process.
Recenty however procedures have also been developed to compute phylogenetic likelihoods under non-reversible models \citep{Boussau2006}, making the use of non-reversible models plausible also for large state-space discrete phytogeographic applications \citep{Edwards2011}.

@@

\subsection{Markov models of codon evolution\label{sub:codon}}
%TODO: briefly introduce, mg is later, show gy94

First two computationally tractable codon substitution models were independently proposed by \cite{Muse1994} and \cite{Goldman1994} in the same issue of Molecular Biology and Evolution (MBE).
In these models, a non-stop codon triplet $n_{1}n_{2}n_{3}$ is considered to be the smallest unit of evolution.
According to the universal genetic code, there are $4^3$ possible triplets minus three stop codons, resulting in a state space size of $61$ codons.
The standard assumption of independence is used, i.e. the substitutions at these three codon positions occur independently, and only a single change per triplet can occur at a given time. 
%%GB: explain in more detail, i.e. more than 1 change is possible in subsequent infinitesimal time intervals
Proteins are coded by a set of 20 amino acids, with each amino acid being coded by a codon triplet. 
Because there are 61 non-stop codons to encode for 20 amino acids, some amino acids will inevitably be coded by more than 1 codon.
A substitution that does not change the encoding amino-acid is called a synonymous substitution, while a non-synonymous substitution refers to a substitution that does result in an amino acid substitution.
Quantifying the rate at which these substitutions occur allows us to infer the selective pressure, which is the main driving force behind molecular evolution.

@@

\subsection{Dated phylogenetic trees\label{sub:dated_trees}}
%how to accommodate time-stamped sequences

Time-stamped sequence data, which frequently arises from studies involving MEP populations for which sequences are sampled at different points in time and the evolutionary rates are high enough for significant changes to occur, can be used to calibrate the molecular clock and convert branch lengths into actual time units.

@@

Central to the Bayesian phylogeographic framework is the hypothesis that sequence evolution is occurring simultaneously with the process of geographical dispersal. 
The observed data, residing at the $N$ tips of phylogeny $\mathbf{F}$, is recorded as both character sequence data $\mathbf{X}=(X_{1},...,X_{N})$ and discretized geographic locations (countries, districts, cities) $\mathbf{Y}=(Y_{1},...,Y_{N})$, as pictured in the upper facette of Figure~\ref{fig:discrete_phylo_illust}.
We assume independent stochastic processes are responsible for generating these data, with ancestral states of $\mathbf{X}$ generated as before by a CTMC characterized by rate matrix $\mathbf{Q}$ and the unobserved ancestral locations $(Y_{N+1},...,Y_{2N-2})$ generated by a separate CTMC characterized by rate matrix $\mathbf{\Phi}$ (see Figure~\ref{fig:discrete_phylo_illust}, lower facette).
The entries of this matrix are the transition rates $\phi_{ij}$ between locations $i$ and $j$.
Unlike before, the rate matrix does not necessarily have to be symmetrical, i.e. there exist some $i,j$ such that $\phi_{ij}\neq\phi_{ji}$.
Additionally, a priori one would expect many of the infinitesimal rates to be zero.

>>>>

As $t$ goes to infinity the probability that the site is in some state $j$ is non-zero and does not depend on the starting state $i$.
This implies that the chain will always display the same long-term behavior, which is why $\mathbf{\Pi}=\{\pi_{i},\ i\in\mathcal{E}\}$ is called a \emph{stationary distribution} (also referred to as the \emph{equilibrium distribution} or \emph{equilibrium frequencies}).
We can look at the elements of $\mathbf{\Pi}$ as the proportion of time spent in state $j\in\mathcal{E}$ after the Markov chain has run for infinitely long time, that does not depend upon the initial state $i$ of the process.
In other words, if we sample the initial state from the stationary distribution and then let the process run for some sufficiently long time $t$, then the distribution of the final state will equal the stationary distribution. 
\myedit{svsixone}{
From the biological perspective stationarity means that the probability distribution of a certain state (e.g. nucleotide character) is constant over the whole course of evolution; the probability to draw a certain state is the same whenever the sampling is done.
A process can be stationary and not be homogeneous, for example by varying substitution rates on the tree branches (see \cite{Yang2002} and the discussion in Subsection~\ref{sub:dpp}).
However the converse is not true, and non-stationarity induces non-homogeneity, as the stochastic process of evolution depends on the stationary distribution.
}

@@

\myedit{svsixtwo}{
Realistic substitution processes are however unlikely to be time-reversible, especially since abandoning stationarity and homogeneity assumptions implies abandoning reversibility of the evolutionary process.
Recenty however procedures have also been developed to compute phylogenetic likelihoods under non-reversible models \citep{Boussau2006}, making the use of non-reversible models plausible also for large state-space discrete phytogeographic applications \citep{Edwards2011}.
}

@@

\subsection{Markov models of codon evolution\label{sub:codon}}

\myedit{svonea}{
First two computationally tractable codon substitution models were independently proposed by \cite{Muse1994} and \cite{Goldman1994} in the same issue of Molecular Biology and Evolution (MBE).
In these models, a non-stop codon triplet $n_{1}n_{2}n_{3}$ is considered to be the smallest unit of evolution.
According to the universal genetic code, there are $4^3$ possible triplets minus three stop codons, resulting in a state space size of $61$ codons.
The standard assumption of independence is used, i.e. the substitutions at these three codon positions occur independently, and only a single change per triplet can occur at a given time. 
%%GB: explain in more detail, i.e. more than 1 change is possible in subsequent infinitesimal time intervals
Proteins are coded by a set of 20 amino acids, with each amino acid being coded by a codon triplet. 
Because there are 61 non-stop codons to encode for 20 amino acids, some amino acids will inevitably be coded by more than 1 codon.
A substitution that does not change the encoding amino-acid is called a synonymous substitution, while a non-synonymous substitution refers to a substitution that does result in an amino acid substitution.
Quantifying the rate at which these substitutions occur allows us to infer the selective pressure, which is the main driving force behind molecular evolution.
}

@@

\subsection{Dated phylogenetic trees\label{sub:dated_trees}}

\myedit{svoneb}{
Time-stamped sequence data, which frequently arises from studies involving MEP populations for which sequences are sampled at different points in time and the evolutionary rates are high enough for significant changes to occur, can be used to calibrate the molecular clock and convert branch lengths into actual time units.
}

@@

\myedit{svonec}{
Central to the Bayesian phylogeographic framework is the hypothesis that sequence evolution is occurring simultaneously with the process of geographical dispersal. 
The observed data, residing at the $N$ tips of phylogeny $\mathbf{F}$, is recorded as both character sequence data $\mathbf{X}=(X_{1},...,X_{N})$ and discretized geographic locations (countries, districts, cities) $\mathbf{Y}=(Y_{1},...,Y_{N})$, as pictured in the upper facette of Figure~\ref{fig:discrete_phylo_illust}.
We assume independent stochastic processes are responsible for generating these data, with ancestral states of $\mathbf{X}$ generated as before by a CTMC characterized by rate matrix $\mathbf{Q}$ and the unobserved ancestral locations $(Y_{N+1},...,Y_{2N-2})$ generated by a separate CTMC characterized by rate matrix $\mathbf{\Phi}$ (see Figure~\ref{fig:discrete_phylo_illust}, lower facette).
The entries of this matrix are the transition rates $\phi_{ij}$ between locations $i$ and $j$.
Unlike before, the rate matrix does not necessarily have to be symmetrical, i.e. there exist some $i,j$ such that $\phi_{ij}\neq\phi_{ji}$.
Additionally, a priori one would expect many of the infinitesimal rates to be zero. 
}
